---
title: "R Notebook"
author: "Carina Huerta"
output:
  html_document:
    df_print: paged
---

```{r}
library(nycflights13)
library(tidyverse)
```
#5.5 Add new variables with mutate()

flights_sml <- select(flights, 
  year:day, 
  ends_with("delay"), 
  distance, 
  air_time
)
mutate(flights_sml,
  gain = dep_delay - arr_delay,
  speed = distance / air_time * 60
)

- mutate() always adds new columns at the end of your dataset so we’ll start by creating a narrower dataset so we can see the new variables

mutate(flights_sml,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)

##5.5.1 Useful creation functions

- Arithmetic operators: +, -, *, /, ^
- Modular arithmetic: %/% (integer division) and %% (remainder), where x == y * (x %/% y) + (x %% y)


transmute(flights,
  dep_time,
  hour = dep_time %/% 100,
  minute = dep_time %% 100
)

- Logs: log(), log2(), log10()
- Offsets: lead() and lag() allow you to refer to leading or lagging values
- Cumulative and rolling aggregates: R provides functions for running sums, products, mins and maxes: cumsum(), cumprod(), cummin(), cummax(); and dplyr provides cummean() for cumulative means
- Logical comparisons, <, <=, >, >=, !=, and ==, which you learned about earlier
- Ranking: there are a number of ranking functions, but you should start with min_rank()
- If min_rank() doesn’t do what you need, look at the variants row_number(), dense_rank(), percent_rank(), cume_dist(), ntile()

```{r}
flights
```

#### 5.5.2 Exercises

1).
```{r}
1504 %/% 100
1504 %% 100
1504 %/% 100 * 60 + 1504 %% 100
```

flights_times <- mutate(flights,
  dep_time_mins = (dep_time %/% 100 * 60 + dep_time %% 100) %% 1440,
  sched_dep_time_mins = (sched_dep_time %/% 100 * 60 +
    sched_dep_time %% 100) %% 1440
)
# view only relevant columns
select(
  flights_times, dep_time, dep_time_mins, sched_dep_time,
  sched_dep_time_mins
)


We could define a function time2mins(), which converts a vector of times in from the format used in flights to minutes since midnight.

2). 

air_time is the difference between the arrival (arr_time) and departure times (dep_time). In other words, air_time = arr_time - dep_time.

- The flight passes midnight, so arr_time < dep_time. In these cases, the difference in airtime should be by 24 hours (1,440 minutes).

- The flight crosses time zones, and the total air time will be off by hours (multiples of 60). 

- Given the time-zones in the US, the differences due to time-zone should be 60 minutes (Central) 120 minutes (Mountain), 180 minutes (Pacific), 240 minutes (Alaska), or 300 minutes (Hawaii).

3). 
the departure delay (dep_delay) to be equal to the difference between scheduled departure time (sched_dep_time), and actual departure time (dep_time), dep_time - sched_dep_time = dep_delay.

4).
```{r}
flights_delayed <- mutate(flights,
  dep_delay_min_rank = min_rank(desc(dep_delay)),
  dep_delay_row_number = row_number(desc(dep_delay)),
  dep_delay_dense_rank = dense_rank(desc(dep_delay))
)
flights_delayed <- filter(
  flights_delayed,
  !(dep_delay_min_rank > 10 | dep_delay_row_number > 10 |
    dep_delay_dense_rank > 10)
)
flights_delayed <- arrange(flights_delayed, dep_delay_min_rank)
print(select(
  flights_delayed, month, day, carrier, flight, dep_delay,
  dep_delay_min_rank, dep_delay_row_number, dep_delay_dense_rank
),
n = Inf
)

```


- dplyr package provides multiple functions for ranking, which differ in how they handle tied values: row_number(), min_rank(), dense_rank()

- row_number() assigns each element a unique value

- min_rank() and dense_rank() assign tied values the same rank, but differ in how they assign values to the next rank

5). 

```{r}
c(1 + 1, 2 + 2, 3 + 3, 1 + 4, 2 + 5, 3 + 6, 1 + 7, 2 + 8, 3 + 9, 1 + 10)
```
6).

- R provides functions for the three primary trigonometric functions: sine (sin())), cosine (cos()), and tangent (tan())


# 5.6 Grouped summaries with summarise()

- The last key verb is summarise(). It collapses a data frame to a single row


##5.6.1 Combining multiple operations with the pipe



by_dest <- group_by(flights, dest)
delay <- summarise(by_dest,
  count = n(),
  dist = mean(distance, na.rm = TRUE),
  delay = mean(arr_delay, na.rm = TRUE)
)
delay <- filter(delay, count > 20, dest != "HNL")

ggplot(data = delay, mapping = aes(x = dist, y = delay)) +
  geom_point(aes(size = count), alpha = 1/3) +
  geom_smooth(se = FALSE)



There are three steps to prepare this data:

- Group flights by destination.

- Summarise to compute distance, average delay, and number of flights.

- Filter to remove noisy points and Honolulu airport, which is almost twice as far away as the next closest airport.

## 5.6.2 Missing values


flights %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay))

-  all aggregation functions have an na.rm argument which removes the missing values prior to computation

## 5.6.3 Counts

delays <- not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(
    delay = mean(arr_delay)
  )

ggplot(data = delays, mapping = aes(x = delay)) + 
  geom_freqpoly(binwidth = 10)


delays <- not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )

ggplot(data = delays, mapping = aes(x = n, y = delay)) + 
  geom_point(alpha = 1/10)


## 5.6.4 Useful summary functions


- Measures of location: we’ve used mean(x), but median(x) is also useful.
- The mean is the sum divided by the length; the median is a value where 50% of x is above it, and 50% is below it.

- Measures of spread: sd(x), IQR(x), mad(x). 
- The root mean squared deviation, or standard deviation sd(x), is the standard measure of spread. 
- The interquartile range IQR(x) and median absolute deviation mad(x) are robust equivalents that may be more useful if you have outliers.

- Measures of rank: min(x), quantile(x, 0.25), max(x). Quantiles are a generalisation of the median. 
      - For example, quantile(x, 0.25) will find a value of x that is greater than 25% of the values, and less than the remaining 75%.

- Measures of position: first(x), nth(x, 2), last(x). 
    - These work similarly to x[1], x[2], and x[length(x)] but let you set a default value if that position does not exist (i.e. you’re trying to get the 3rd element from a group that only has two elements).

- Counts: You’ve seen n(), which takes no arguments, and returns the size of the current group. 
- To count the number of non-missing values, use sum(!is.na(x)).
- To count the number of distinct (unique) values, use n_distinct(x)

- Counts and proportions of logical values: sum(x > 10), mean(y == 0). 
- When used with numeric functions, TRUE is converted to 1 and FALSE to 0. 
- This makes sum() and mean() very useful: sum(x) gives the number of TRUEs in x, and mean(x) gives the proportion.


## 5.6.5 Grouping by multiple variables

daily <- group_by(flights, year, month, day)
(per_day   <- summarise(daily, flights = n()))


(per_month <- summarise(per_day, flights = sum(flights)))

(per_year  <- summarise(per_month, flights = sum(flights)))

## 5.6.6 Ungrouping

- to remove grouping, and return to operations on ungrouped data, use ungroup(

## 5.6.7 Exercises


1). If a flight is always 30 minutes late and that delay is known, then it is as if the arrival time is that delayed time.


2).
not_cancelled %>%
  group_by(dest) %>%
  summarise(n = length(dest))

3).

filter(flights, !is.na(dep_delay), is.na(arr_delay)) %>%
  select(dep_time, arr_time, sched_arr_time, dep_delay, arr_delay)
- most important column is arr_delay, which indicates the amount of delay in arrival.

4).
cancelled_and_delays <-
  flights %>%
  mutate(cancelled = (is.na(arr_delay) | is.na(dep_delay))) %>%
  group_by(year, month, day) %>%
  summarise(
    cancelled_prop = mean(cancelled),
    avg_dep_delay = mean(dep_delay, na.rm = TRUE),
    avg_arr_delay = mean(arr_delay, na.rm = TRUE)
  ) %>%

ggplot(cancelled_and_delays) +
  geom_point(aes(x = avg_arr_delay, y = cancelled_prop))
  
  
  5).
  
  
```{r}
flights %>%
  group_by(carrier) %>%
  summarise(arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
  arrange(desc(arr_delay))
```
  
6). The sort argument to count() sorts the results in order of n. You could use this anytime you would run count() followed by arrange()


# 5.7.1 Exercises


1).

- Summary functions (mean()), offset functions (lead(), lag()), ranking functions (min_rank(), row_number()), operate within each group when used with group_by() in mutate() or filter().
- Arithmetic operators (+, -), logical operators (<, ==), modular arithmetic operators (%%, %/%), logarithmic functions (log) are not affected by group_by.



2).

```{r eval=FALSE}
flights %>%
  filter(!is.na(tailnum)) %>%
  mutate(on_time = !is.na(arr_time) & (arr_delay <= 0)) %>%
  group_by(tailnum) %>%
  summarise(on_time = mean(on_time), n = n()) %>%
  filter(min_rank(on_time) == 1)

```
3). 
```{r}
flights %>%
  group_by(hour) %>%
  summarise(arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
  arrange(arr_delay)
```
4).
```{r}
cancelled_and_delays <-
  flights %>%
  mutate(cancelled = (is.na(arr_delay) | is.na(dep_delay))) %>%
  group_by(year, month, day) %>%
  summarise(
    cancelled_prop = mean(cancelled),
    avg_dep_delay = mean(dep_delay, na.rm = TRUE),
    avg_arr_delay = mean(arr_delay, na.rm = TRUE)
  ) %>%
  ungroup()
```

```{r}
flights %>%
  filter(arr_delay > 0) %>%
  group_by(dest) %>%
  mutate(
    arr_delay_total = sum(arr_delay),
    arr_delay_prop = arr_delay / arr_delay_total
  ) %>%
  select(
    dest, month, day, dep_time, carrier, flight,
    arr_delay, arr_delay_prop
  ) %>%
  arrange(dest, desc(arr_delay_prop))

```

5).  After about 8-hours, a delayed flight is likely to be followed by a flight leaving on time.

```{r}
lagged_delays <- flights %>%
  arrange(origin, month, day, dep_time) %>%
  group_by(origin) %>%
  mutate(dep_delay_lag = lag(dep_delay)) %>%
  filter(!is.na(dep_delay), !is.na(dep_delay_lag))
```

```{r eval=FALSE}
lagged_delays %>%
  group_by(dep_delay_lag) %>%
  summarise(dep_delay_mean = mean(dep_delay)) %>%
  ggplot(aes(y = dep_delay_mean, x = dep_delay_lag)) +
  geom_point() +
  scale_x_continuous(breaks = seq(0, 1500, by = 120)) +
  labs(y = "Departure Delay", x = "Previous Departure Delay")
```

```{r eval=FALSE}
lagged_delays %>%
  group_by(origin, dep_delay_lag) %>%
  summarise(dep_delay_mean = mean(dep_delay)) %>%
  ggplot(aes(y = dep_delay_mean, x = dep_delay_lag)) +
  geom_point() +
  facet_wrap(~origin, ncol = 1) +
  labs(y = "Departure Delay", x = "Previous Departure Delay")
```

6)
```{r}
flights %>%
  mutate(mph = distance / (air_time / 60)) %>%
  arrange(desc(mph)) %>%
  select(
    origin, dest, mph, year, month, day, dep_time, flight, carrier,
    dep_delay, arr_delay
  )

```

7).

```{r}
flights %>%
  # find all airports with > 1 carrier
  group_by(dest) %>%
  mutate(n_carriers = n_distinct(carrier)) %>%
  filter(n_carriers > 1) %>%
  # rank carriers by numer of destinations
  group_by(carrier) %>%
  summarize(n_dest = n_distinct(dest)) %>%
  arrange(desc(n_dest))

```


```{r}
filter(airlines, carrier == "EV")
filter(airlines, carrier %in% c("AS", "F9", "HA"))
```

8).The exception is flights on the days on which daylight savings started (March 10) or ended (November 3). 

```{r}
flights %>%
  select(tailnum, year, month, day, dep_delay) %>%
  filter(!is.na(dep_delay)) %>%
  arrange(tailnum, year, month, day) %>%
  group_by(tailnum) %>%
  mutate(cumulative_hr_delays = cumsum(dep_delay > 60)) %>%
  summarise(total_flights = sum(cumulative_hr_delays < 1)) %>%
  arrange(total_flights)

```






